---
title: "Visualize Balance Sheets"
author: "Darko Bergant"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualize Balance Sheets}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo=FALSE, results='hide', message=FALSE }
library(dplyr)
library(tidyr)
library(finstr)
data(xbrl_data_aapl2013)
data(xbrl_data_aapl2014)
```

**Warning: finstr package is in development. 
Please use with caution.**

## Get data
Use XBRL package to parse XBRL files. For example:
```{r xbrl_parse_min, eval=FALSE, echo=TRUE}
library(XBRL)
# parse XBRL (Apple 10-K report)
xbrl_url2014 <- 
  "http://edgar.sec.gov/Archives/edgar/data/320193/000119312514383437/aapl-20140927.xml"
xbrl_url2013 <- 
  "http://edgar.sec.gov/Archives/edgar/data/320193/000119312513416534/aapl-20130928.xml"
xbrl_data_aapl2014 <- xbrlDoAll(xbrl_url2014)
xbrl_data_aapl2013 <- xbrlDoAll(xbrl_url2013)
```

## Prepare statements
With `xbrl_get_statements` convert XBRL data to *statements* object. 
```{r xbrl_get_statements}
library(finstr)

st2013 <- xbrl_get_statements(xbrl_data_aapl2013)
st2014 <- xbrl_get_statements(xbrl_data_aapl2014)
# merge all statements
st_all <- merge( st2013, st2014 )
# get balance sheets 
balance_sheet <- st_all$StatementOfFinancialPositionClassified
tail(balance_sheet,2)

```

## Prepare custom calculation hierarchy
There is no human readable way to plot every number of the balance sheet in 
one graph.
The only way to plot a balance sheet is to plot it several times. 
Each graph should have a limited number of highlited features.
The first step is to break a balance sheet to a small number of pieces.
We can use calculations to specify these groups of elements.

```{r custom_hierarchy}

two_sided_bs_calculation <- 
  list(
    "Assets" = calculation(
      "Cash and Equivalents" = CashAndCashEquivalentsAtCarryingValue,
      "Other Current Assets" = AssetsCurrent - CashAndCashEquivalentsAtCarryingValue,
      "Other Assets" = Assets - AssetsCurrent
    ),
    
    "Liabilities and Equity" = calculation(
      "Current Liabilities" = LiabilitiesCurrent,
      "Other Liabilities" =  Liabilities - LiabilitiesCurrent,
      "Stockholders Equity" = StockholdersEquity
    )
  )

```

We divided balance sheet to **Assets** and **Liabilities and Equity**. 
Both main groups are divided to only 3 smaller chunks (based on liquidity). 

To plot the result we need to run the calculations on a balance sheet and call
graph plotting function:

```{r graph1, fig.width=7.3}

balance_sheet %>% 
  do_calculation(two_sided_bs_calculation) %>%
  plot_double_stacked_bar()

```

Another option is to group by date and see assets close to liabilities for
every year:

```{r graph2, fig.width=7.3}

balance_sheet %>% 
  do_calculation(two_sided_bs_calculation) %>%
  plot_double_stacked_bar(by_date = TRUE)

```

## See the difference
We can use the same custom hierarchy on lagged differences.

```{r graph3, fig.width=7.3}

balance_sheet %>%
  diff() %>%
  do_calculation(two_sided_bs_calculation) %>%
  plot_double_stacked_bar(
    by_date = TRUE, is_diff = TRUE, 
    dif_labels = c("Money\nconsumption","Money\nsupply"))

```








